---
title: "SIMP59 Lab 1: Text Tokenization and Sentiment Analysis"
output: html_document
---

## Text Tokenization

### 1) Setup

This section reads `.txt` files from `text-data/` and builds three tidy data frames:
`df_texts`, `df_sentences`, and `df_tokens`.

```{r p1_setup, message=FALSE, warning=FALSE}
required_pkgs <- c(
  "dplyr", "tidyr", "readr", "purrr", "stringr",
  "tidytext", "ggplot2", "textstem", "wordcloud", "RColorBrewer"
)

missing_pkgs <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))
]

if (length(missing_pkgs) > 0) {
  stop(
    paste("Missing required packages:", paste(missing_pkgs, collapse = ", ")),
    call. = FALSE
  )
}

library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(stringr)
library(tidytext)
library(ggplot2)
library(textstem)

dir_input <- "text-data"
dir_output <- "output"
dir.create(dir_output, showWarnings = FALSE, recursive = TRUE)

txt_files <- list.files(
  path = dir_input,
  pattern = "\\.txt$",
  full.names = TRUE
)

if (length(txt_files) == 0) {
  stop("No .txt files found in text-data/.", call. = FALSE)
}
```

### 2) Import Raw Text Files to `df_texts`

Each file becomes one row with `filename` and full `text_data`.

```{r p1_df_texts}
df_texts <- tibble(file_path = txt_files) %>%
  mutate(
    filename = basename(file_path),
    text_data = map_chr(file_path, readr::read_file)
  ) %>%
  select(filename, text_data)

df_texts %>% slice_head(n = 5)
```

### 3) Split into Sentences: `df_sentences`

Split each document into sentences and number them per file.

```{r p1_df_sentences}
df_sentences <- df_texts %>%
  unnest_tokens(
    output = sentence_text,
    input = text_data,
    token = "sentences",
    to_lower = FALSE
  ) %>%
  group_by(filename) %>%
  mutate(sentence_number = row_number()) %>%
  ungroup() %>%
  select(filename, sentence_number, sentence_text)

df_sentences %>% slice_head(n = 10)
```

### 4) Tokenize and Normalize: `df_tokens`

Tokens are lowercased, stripped to alphabetic terms, stopwords are removed,
then lemmatized to base forms with `textstem`.

```{r p1_df_tokens}
token_normalizer <- function(x) {
  textstem::lemmatize_words(x)
}

df_tokens <- df_sentences %>%
  unnest_tokens(
    output = token_text,
    input = sentence_text,
    token = "words",
    to_lower = TRUE
  ) %>%
  filter(str_detect(token_text, "^[a-z]+$")) %>%
  anti_join(tidytext::stop_words, by = c("token_text" = "word")) %>%
  mutate(token_text = token_normalizer(token_text)) %>%
  filter(str_detect(token_text, "^[a-z]+$"), nchar(token_text) > 2) %>%
  group_by(filename, sentence_number) %>%
  mutate(token_number = row_number()) %>%
  ungroup() %>%
  select(filename, sentence_number, token_number, token_text)

df_tokens %>% slice_head(n = 15)
```

### 5) Save Data Frames as TSV

Write each tidy data frame to disk as a tab-separated file.

```{r p1_save_tsv}
readr::write_tsv(df_texts, file.path(dir_output, "df_texts.tsv"))
readr::write_tsv(df_sentences, file.path(dir_output, "df_sentences.tsv"))
readr::write_tsv(df_tokens, file.path(dir_output, "df_tokens.tsv"))

tibble(
  output_file = c("df_texts.tsv", "df_sentences.tsv", "df_tokens.tsv"),
  path = file.path(dir_output, output_file)
)
```

### 6) Horizontal Frequency Bar Plot per Filename

Compute token frequency by file and show top content terms.

```{r p1_bar_plot, fig.width=14, fig.height=10}
df_token_freq <- df_tokens %>%
  count(filename, token_text, sort = TRUE)

df_top_tokens <- df_token_freq %>%
  group_by(filename) %>%
  slice_max(n, n = 15, with_ties = FALSE) %>%
  ungroup()

p_freq <- ggplot(
  df_top_tokens,
  aes(
    x = tidytext::reorder_within(token_text, n, filename),
    y = n,
    fill = filename
  )
) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ filename, scales = "free_y") +
  tidytext::scale_x_reordered() +
  labs(
    title = "Top Token Frequencies by Filename",
    x = "Token",
    y = "Frequency"
  )

p_freq

ggsave(
  filename = file.path(dir_output, "token_frequency_by_filename.png"),
  plot = p_freq,
  width = 14,
  height = 10,
  dpi = 300
)
```

### 7) Word Cloud Across All Filenames

Aggregate all tokens and display one global word cloud.

```{r p1_wordcloud_plot, fig.width=12, fig.height=8}
df_wordcloud <- df_tokens %>%
  count(token_text, sort = TRUE)

set.seed(123)
wordcloud::wordcloud(
  words = df_wordcloud$token_text,
  freq = df_wordcloud$n,
  max.words = 200,
  random.order = FALSE,
  colors = RColorBrewer::brewer.pal(8, "Dark2"),
  scale = c(4, 0.8)
)

png(
  filename = file.path(dir_output, "wordcloud_all_filenames.png"),
  width = 1600,
  height = 1000,
  res = 150
)
wordcloud::wordcloud(
  words = df_wordcloud$token_text,
  freq = df_wordcloud$n,
  max.words = 200,
  random.order = FALSE,
  colors = RColorBrewer::brewer.pal(8, "Dark2"),
  scale = c(4, 0.8)
)
dev.off()
```

## Sentiment Analysis

### 1) Setup and Load `df_tokens`

Load the token-level data from `output/df_tokens.tsv`.

```{r p2_setup, message=FALSE, warning=FALSE}
required_pkgs <- c("dplyr", "readr", "tidyr", "ggplot2", "syuzhet")

missing_pkgs <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))
]

if (length(missing_pkgs) > 0) {
  stop(
    paste("Missing required packages:", paste(missing_pkgs, collapse = ", ")),
    call. = FALSE
  )
}

library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(syuzhet)

dir_output <- "output"
tokens_path <- file.path(dir_output, "df_tokens.tsv")

if (!file.exists(tokens_path)) {
  stop("Missing input file: output/df_tokens.tsv", call. = FALSE)
}

df_tokens <- readr::read_tsv(tokens_path, show_col_types = FALSE)

df_tokens %>% slice_head(n = 10)
```

### 2) Load AFINN Dictionary

Use AFINN scores for token-level sentiment values.

```{r p2_afinn_dict}
afinn <- syuzhet::get_sentiment_dictionary("afinn") %>%
  tibble::as_tibble() %>%
  rename(token_text = word, afinn_value = value)

afinn %>% slice_head(n = 10)
```

### 3) Sentiment per Sentence (`sentence_number`)

Join tokens to AFINN and sum sentiment values within each sentence.

```{r p2_sentence_sentiment}
df_sentence_scored <- df_tokens %>%
  inner_join(afinn, by = "token_text")

df_sentence_sentiment <- df_tokens %>%
  distinct(filename, sentence_number) %>%
  left_join(
    df_sentence_scored %>%
      group_by(filename, sentence_number) %>%
      summarize(
        sentence_sentiment = sum(afinn_value),
        scored_tokens = n(),
        .groups = "drop"
      ),
    by = c("filename", "sentence_number")
  ) %>%
  mutate(
    sentence_sentiment = replace_na(sentence_sentiment, 0),
    scored_tokens = replace_na(scored_tokens, 0)
  ) %>%
  arrange(filename, sentence_number)

df_sentence_sentiment %>% slice_head(n = 12)
```

### 4) Visualize Sentence Sentiment per Filename

Plot sentence-level sentiment trajectories for each text file.

```{r p2_sentence_plot, fig.width=13, fig.height=9}
ggplot(df_sentence_sentiment, aes(x = sentence_number, y = sentence_sentiment)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(color = "#1f77b4", linewidth = 0.4) +
  facet_wrap(~ filename, scales = "free_x") +
  labs(
    title = "AFINN Sentiment by Sentence Number",
    x = "Sentence Number",
    y = "Sentence Sentiment (sum of AFINN scores)"
  )
```

### 5) Overall Sentiment per Filename (`df_sentiment`)

Aggregate sentiment by file and build `df_sentiment`.

```{r p2_overall_sentiment}
df_sentiment <- df_sentence_sentiment %>%
  group_by(filename) %>%
  summarize(
    total_sentiment = sum(sentence_sentiment),
    mean_sentence_sentiment = mean(sentence_sentiment),
    positive_sentences = sum(sentence_sentiment > 0),
    negative_sentences = sum(sentence_sentiment < 0),
    neutral_sentences = sum(sentence_sentiment == 0),
    n_sentences = n(),
    .groups = "drop"
  ) %>%
  left_join(
    df_tokens %>%
      count(filename, name = "n_tokens"),
    by = "filename"
  ) %>%
  mutate(sentiment_per_100_tokens = 100 * total_sentiment / n_tokens) %>%
  arrange(desc(total_sentiment))

df_sentiment
```

### 6) Visualize Overall Sentiment per Filename

Compare total sentiment across all text files.

```{r p2_overall_plot, fig.width=10, fig.height=6}
ggplot(
  df_sentiment,
  aes(x = reorder(filename, total_sentiment), y = total_sentiment, fill = total_sentiment > 0)
) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Overall AFINN Sentiment by Filename",
    x = "Filename",
    y = "Total Sentiment"
  )
```

### 7) Save Sentiment Data Frames as TSV

Save sentence-level and file-level sentiment outputs.

```{r p2_save_tsv}
readr::write_tsv(df_sentence_sentiment, file.path(dir_output, "df_sentence_sentiment.tsv"))
readr::write_tsv(df_sentiment, file.path(dir_output, "df_sentiment.tsv"))

tibble::tibble(
  output_file = c("df_sentence_sentiment.tsv", "df_sentiment.tsv"),
  path = file.path(dir_output, output_file)
)
```
