---
title: "WVS Wave 7 Data Analysis Notebook"
output: html_document
---

This notebook documents a complete analysis workflow using a cleaned TSV file loaded from GitHub.
The goal is to inspect country coverage, clean special missing-value codes, and analyze patterns
in life satisfaction by age, country, sex, and confidence in parliament.

## 1. Load Libraries and Data

In the next chunk, we load the required packages and read the TSV file from a public URL that contains only the relevant variables used in this analysis.

```{r setup}
library(dplyr)
library(ggplot2)
library(tidyr)
library(forcats)

data_url <- "https://raw.githubusercontent.com/cca-cce/osm-comm/refs/heads/main/csv/WVS_Cross-National_Wave_7_csv_v6_0.tsv"
wvs_small <- read.delim(data_url, sep = "\t", stringsAsFactors = FALSE, check.names = FALSE)

glimpse(wvs_small)
summary(wvs_small)
```

## 2. Data Cleaning and Feature Engineering

In the next chunk, we convert `country` to a factor, replace negative survey codes with `NA`, and create analysis-friendly variables for age class and sex labels.

```{r clean-data}
wvs_clean <- wvs_small %>%
  mutate(
    country = as.factor(country),
    age = ifelse(age < 0, NA, age),
    sex = ifelse(sex < 0, NA, sex),
    life_sat = ifelse(life_sat < 0, NA, life_sat),
    trust_people = ifelse(trust_people < 0, NA, trust_people),
    conf_parliament = ifelse(conf_parliament < 0, NA, conf_parliament),
    age_class = case_when(
      is.na(age) ~ NA_character_,
      age >= 16 & age <= 25 ~ "Young",
      age >= 55 ~ "Old",
      TRUE ~ "Middle"
    ),
    age_class = factor(age_class, levels = c("Young", "Middle", "Old")),
    sex_label = case_when(
      sex == 1 ~ "Male",
      sex == 2 ~ "Female",
      TRUE ~ NA_character_
    )
  )

summary(wvs_clean)
```

This cleaning step creates a consistent base table for the rest of the notebook, ensuring that missing data does not bias means and correlations.

## 3. Country Coverage

In the next chunk, we count observations per country and visualize the top 20 countries by sample size to understand the distribution of cases.

```{r country-coverage}
country_table <- wvs_clean %>%
  count(country, name = "observations") %>%
  arrange(desc(observations))

head(country_table, 10)

country_table %>%
  slice_head(n = 20) %>%
  ggplot(aes(x = fct_reorder(country, observations), y = observations)) +
  geom_col(fill = "#1f78b4") +
  coord_flip() +
  labs(
    title = "Top 20 Countries by Number of Observations",
    x = "Country",
    y = "Observations"
  ) +
  theme_minimal()
```

The chart shows where most of the sample is concentrated, which is useful context when interpreting global averages.

## 4. Life Satisfaction by Age Class (Global)

In the next chunk, we compute and visualize mean life satisfaction for each age class across the full dataset.

```{r life-sat-global}
life_sat_table <- wvs_clean %>%
  group_by(age_class) %>%
  summarize(mean_life_sat_world = mean(life_sat, na.rm = TRUE), .groups = "drop")

life_sat_table

ggplot(life_sat_table, aes(x = age_class, y = mean_life_sat_world, fill = age_class)) +
  geom_col(width = 0.7, alpha = 0.9, show.legend = FALSE) +
  labs(
    title = "Mean Life Satisfaction by Age Class (Global)",
    x = "Age Class",
    y = "Mean Life Satisfaction"
  ) +
  theme_minimal()
```

This summarizes the global age pattern and gives a baseline before country-specific comparison.

## 5. Global vs USA Comparison

In the next chunk, we calculate USA-specific age-class means, join them with global means, and plot both series together.

```{r global-vs-usa}
usa_life_sat_table <- wvs_clean %>%
  filter(country == "USA") %>%
  group_by(age_class) %>%
  summarize(mean_life_sat_usa = mean(life_sat, na.rm = TRUE), .groups = "drop")

life_sat_comparison <- left_join(life_sat_table, usa_life_sat_table, by = "age_class") %>%
  rename(world = mean_life_sat_world, usa = mean_life_sat_usa)

life_sat_comparison

life_sat_comparison_long <- life_sat_comparison %>%
  pivot_longer(cols = c(world, usa), names_to = "group", values_to = "mean_life_sat")

ggplot(life_sat_comparison_long, aes(x = age_class, y = mean_life_sat, color = group, group = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  labs(
    title = "Life Satisfaction by Age Class: World vs USA",
    x = "Age Class",
    y = "Mean Life Satisfaction",
    color = "Group"
  ) +
  theme_minimal()
```

This comparison highlights whether the USA follows or diverges from the global age-related life satisfaction pattern.

## 6. Gender Differences Across Age Classes

In the next chunk, we estimate mean life satisfaction by age class and sex, then visualize group differences with side-by-side bars.

```{r gender-differences}
life_sat_gender_table <- wvs_clean %>%
  filter(!is.na(sex_label), !is.na(age_class)) %>%
  group_by(age_class, sex_label) %>%
  summarize(mean_life_sat = mean(life_sat, na.rm = TRUE), .groups = "drop")

life_sat_gender_table

ggplot(life_sat_gender_table, aes(x = age_class, y = mean_life_sat, fill = sex_label)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  labs(
    title = "Mean Life Satisfaction by Age Class and Sex",
    x = "Age Class",
    y = "Mean Life Satisfaction",
    fill = "Sex"
  ) +
  theme_minimal()
```

This step extends the age analysis by adding a gender perspective within each age category.

## 7. Correlation Between Life Satisfaction and Confidence in Parliament

In the next chunk, we run a correlation test and create a scatterplot with a fitted linear trend line.

```{r correlation-analysis}
cor_result <- cor.test(wvs_clean$life_sat, wvs_clean$conf_parliament)
cor_result

ggplot(wvs_clean, aes(x = conf_parliament, y = life_sat)) +
  geom_jitter(width = 0.15, height = 0.15, alpha = 0.08, color = "#4d4d4d") +
  geom_smooth(method = "lm", se = FALSE, color = "#e31a1c", linewidth = 1) +
  labs(
    title = "Life Satisfaction vs Confidence in Parliament",
    x = "Confidence in Parliament",
    y = "Life Satisfaction"
  ) +
  theme_minimal()
```

The correlation output and fitted line provide a quick check of the strength and direction of association between political confidence and life satisfaction.

## 8. Analysis Narrative Summary

The notebook follows a standard analysis pipeline:
1. Load a compact TSV containing only relevant variables.
2. Clean negative response codes and derive age/sex categories.
3. Describe sample structure by country.
4. Estimate global life-satisfaction patterns by age.
5. Compare global results with USA-specific estimates.
6. Add gender-based differences within age classes.
7. Evaluate linear association between life satisfaction and confidence in parliament.

These steps produce an interpretable and reproducible workflow that can be extended for deeper country-level or subgroup analysis.
